Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-25 WSSSPE1_report/WSSSPE1_summary_report.tex
adding comments and making small changes in 4, partly done but stopping for now

The panel on developing and supporting software examined the challenges
around scientific software development/support mainly focused on research
groups that also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several stated that documentation is not just for users, and paying attention
to API documentation, tutorials for building and deploying software, along with
documented development practices can be very helpful in bringing  new developers
into a project efficiently.

There was discussion that backward compatibility is not always desirable, and it
can be very costly. This must be balanced with the aims of a given project, how
many other projects depend on and use the code when backwards incompatible
changes are to be made. There are many examples in the wider open source
software world of strategies for dealing with this, and again best practices
can go a long way to mitigating issues around backwards compatibility. Many
projects live with sub-optimal code for a while, and make allow backwards
compatibility to be broken at agreed upon development points, such as a major
release for a software library.

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the papers submitted discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as lists where developers do not participate in
mailing lists as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek
and Turk's paper on scaling code in the human dimension discuss how development
lists that are welcoming to people actually have many more people contributing.

Discussion around recruiting and/or retaining personnel in this area is hard,
with one of the major reasons being no long-term career paths (especially when
compared with industry). How should software development be balanced with
research? It is apparent that things are beginning to improve, such as
incentives for software development in the form of altmetrics, tenure committee
consideration, NSF Products vs Publications is was noted it can be very
difficult to measure where people are using small bits of your code.

There were 13 articles about different experiences in this area, but little
about GUI testing, performance, scalability, or agile development practices.
There were several unique perspectives about issues such as managing API
changes, using the same best practices for software as data, and going beyond
simply ``slapping an OSI-approved license on code''.

The question of what ``sustainable'' means in the context of software was
raised. The resources at http://oss-watch.ac.uk/resources/ssmm discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Questions such as whether the supplier will still be around in five
years time, will it still care for its customers needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

@@

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Everyone thinks that you can't rewrite your software, but if it is
modular then you can keep systems up to date. Extensibility will keep you
relevant if it is built into the project. One counterpoint raised by Jason Riedy
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware''.

>>>>

The panel on developing and supporting software examined the challenges
around scientific software development/support, mainly focused on research
groups that also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several stated that documentation is not just for users, and paying attention
to API documentation, tutorials for building and deploying software, along with
documented development practices can be very helpful in bringing  new developers
into a project efficiently.

There was discussion that backward compatibility is not always desirable, and it
can be very costly. This must be balanced with the aims of a given project, and how
many other projects depend on and use the code when backwards incompatible
changes are to be made. There are many examples in the wider open source
software world of strategies for dealing with this, and again best practices
can go a long way to mitigating issues around backwards compatibility. Many
projects live with sub-optimal code for a while, and may allow backwards
compatibility to be broken at agreed upon development points, such as a major
release for a software library.

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as lists where developers do not participate in
mailing lists as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
lists that are welcoming to people actually have many more people contributing.

Recruiting and/or retaining personnel in this area is hard,
with one of the major reasons being no long-term career paths (especially when
compared with industry). How should software development be balanced with
research? It is apparent that things are beginning to improve, such as
incentives for software development in the form of altmetrics, tenure committee
consideration, and NSF ``products'' vs ``publications''. It was noted it can be very
difficult to measure where people are using small bits of your code.

There were 13 articles about different experiences in this area, but little
about GUI testing, performance, scalability, or agile development practices.
There were several unique perspectives about issues such as managing API
changes, using the same best practices for software as data, and going beyond
simply ``slapping an OSI-approved license on code.''

The question of what ``sustainable'' means in the context of software was
raised (see \S\ref{sec:cross-cutting}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used,
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Important questions include whether the supplier will still be around in five
years time, will it still care for its customers needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

@@

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Everyone \note{everyone in the audience?  paper authors?  software developers in general?} thinks that you can't rewrite your software, but if it is
modular then you can keep systems up to date. Extensibility will keep you
relevant if it is built into the project. One counterpoint raised by Jason Riedy\todo{add cite}
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware''.
